---
title: "Gov 50 Final Project"
author: Jack Wyss
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---

```{r setup}
library(broom)
library(infer)
library(moderndive)
library(skimr)
library(tidyverse)
```



## 1. Introduction

How do movie sequels compare to the original, in terms of box office return and audience reception? Does a better original film make for better sequels, or worse? Many people criticize the production of movie sequels, and yet studios keep making them. I hypothesize that better-received films will produce better-received and more profitable sequels. These films might continue the story of the franchise and be remembered just as well as the original, whereas the sequel to a poorly-received comedy might also be poorly-received, both critically and by people buying tickets. This data is worth analyzing because movies and sequels are a continually relevant part of global cultural consumption and inform how we think about politics, news, and the arts.

If box office return of the original and average box office return of sequels are positively correlated, then there is evidence for part of my hypothesis. If the IMDb score of the original and the average IMDb score of its sequels are positively correlated, there is evidence for the other part of my hypothesis. If there is a negative correlation or no correlation for either, my hypothesis will be weakened. I will focus only on the first, second, and third films in a series for convenience, but there are still interesting things to be discovered in the data despite this limitation.

## 2. Data

There are 869 movies in this data set. The data notes what franchise a film belongs to and the order in which it was released. For example, the original "Halloween" movie is given a value in the "Order" column of 1. The explanatory variables in this case are the box office revenue of the original, measured in dollars, and the IMDb score of the original, measured on a scale of 1 to 10 (with 10 being the best). The response variables are the box office revenue and IMDb scores of the original film's sequels. Since some sequels might be better-received than others, I will take the average gross box office return and IMDb score of the entire franchise after the first film. **Note that for movies released before IMDb existed the score is retroactive and may reflect modern consumer tastes rather than the tastes of people who went to see the movie in theaters.**


```{r}

movies <- read_csv("movies_metadata.csv")|>
  select(budget, overview, title, revenue, release_date, runtime)|>
  mutate(release_date = as.character(release_date))

sequels <- read_csv("Sequels/priyankad0993-sequels/Movies - Cleaned.csv")

sequels|>
  filter(`Movie Series` == "A Nightmare on Elm Street")

movies_sequels <- sequels|>
  left_join(movies, by = c("Title" = "title", "Runtime (mins)" = "runtime"))|>
    select(Title, `Movie Series`, `IMDb Rating`, Order, budget, overview, revenue, `Runtime (mins)`)|>
  mutate(budget = as.integer(budget),
         revenue = as.integer(revenue),
    Order = as.character(Order))|>
  rename(series = `Movie Series`,
         rating = `IMDb Rating`)

movies_sequels|>filter(series == "A Nightmare on Elm Street")

```

## 3. Analysis

**a) Average Number of Movies in a Series**


In this data set, movies are grouped by series, and each series has a certain number of movies. Some series have many sequels, and others are "one-and-done," so to speak; there is only one movie in the series. The average number of movies in series that have sequels is about 2.88. You can see on the histogram below that many series have two or three movies, whereas some have 10 or more. For the purposes of this project, I will be focusing on the first, second, and third films in a series.

```{r}

numbermovies <- movies_sequels|>
  group_by(series)|>
  summarize(number_of_movies = n())|>
  filter(number_of_movies > 1)

  mean(numbermovies$number_of_movies)
```
```{r}

numbermovies_plot <- numbermovies|>
  ggplot(mapping = aes(x = number_of_movies))+
  geom_bar()+
  labs(
    title = "Number of Movies by Franchise",
    x = "Number of Movies in the Franchise",
    y = "Count",
    caption = "Note: Excludes franchises with only one movie in the dataset."
  )+
  geom_vline(xintercept = mean(numbermovies$number_of_movies), color = "purple", size = 1.25)

numbermovies_plot

```

**b) Box Office Return (General)**

There is a strong positive correlation (0.7784) between a movie's budget and its box office return. As a movie's budget increases, it will generally make more money at the box office. Note that I filtered both the budget and revenue of a film to be greater than $10,000 to avoid capturing movies that did not get a theatrical release, or whose budgets are unknown.

```{r}

revenueplot <- movies_sequels|>
  drop_na()|>
  filter(budget > 10000 | revenue > 10000)|>
  ggplot(mapping = aes(x = budget, y = revenue, color = Order))+
  geom_point()+
  geom_smooth(method = "lm", se = FALSE, color = "purple")+
  labs(title = "Budget vs Box Office Revenue",
       x = "Budget",
       y = "Revenue"
       )+
  scale_x_log10()+
  scale_y_log10()
  
revenueplot

cor(movies_sequels$budget, movies_sequels$revenue, use = "pairwise")
```

```{r}


movies_sequels <- movies_sequels|>
  mutate(sequel_category = case_when(
    Order == 1 ~ "First",
    Order == 2 ~ "Second",
    Order == 3 ~ "Third",
    Order == 4 ~ "Fourth",
    Order == 5 ~ "Fifth",
    Order == 6 ~ "Sixth",
    Order == 7 ~ "Seventh",
    Order == 8 ~ "Eighth",
    Order >= 9 ~ "Ninth +"),
    is.sequel = if_else(Order == 1, "No", "Yes")
  )

has_sequels_tibble <- movies_sequels|>
  group_by(series)|>
  summarize(number_of_movies = n())|>
  mutate(series.has.sequels = if_else(number_of_movies > 1, "Yes", "No"))

movies_sequels_new <- movies_sequels|>
inner_join(has_sequels_tibble, by = "series")

```

**c) Box Office Return for Sequels vs Original**

The box-office revenue of sequels in a series has a strong positive correlation (r = 0.7753) with the box-office revenue of the first film in the series: as the revenue of the first film increases, so, too, does the sequels' revenue. The slope of the least-squares regression line is 0.9, meaning that when the revenue of an original or first film increases by one dollar, the sequel revenue will increase by about $0.90. This result is statistically significant because p < 0.05. According to the r-squared value, about 60.1% of the variation in the data is explained by this model. However, I don't think there is a causal relationship between the first film's revenue and the sequel revenue -- in other words, a lot of revenue for the original film does not *cause* the sequels to have higher revenue. Perhaps there are other factors (e.g. critical reception) that influence the sequels' reception.

```{r}

firstfilm <- movies_sequels_new|>
  drop_na()|>
  filter(series.has.sequels == "Yes")|>
  group_by(series, Order)|>
  summarize(first_film_revenue = revenue,
            first_film_imdb = rating)|>
  filter(Order == 1)

sequelfilms <- movies_sequels_new|>
  drop_na()|>
  filter(series.has.sequels == "Yes")|>
  group_by(series, Order)|>
  summarize(sequels_revenue = revenue,
            sequels_imdb = rating)|>
  filter(Order == 2 | Order == 3)

first_second_third <- firstfilm|>
  inner_join(sequelfilms, by = c("series"))|>
  rename(Order = Order.y)|>
  filter(first_film_revenue > 10000 | sequels_revenue > 10000)

first_second_third

sequel_revenue_plot <- first_second_third|>
  ggplot(first_second, mapping = aes(x = first_film_revenue, y = sequels_revenue, color = Order))+
  geom_point()+
  geom_smooth(method = "lm", se = FALSE)+
  labs(
    title = "Revenue of Sequels vs First Film Revenue",
    x = "First Film Revenue",
    y = "Sequel Revenue"
  )

sequel_revenue_plot

```

```{r}

lm_fst <- lm(sequels_revenue ~ first_film_revenue, data = first_second_third)

cor(first_second_third$first_film_revenue, first_second_third$sequels_revenue)

glance_lm <- glance(lm_fst)|>
  select(r.squared)

knitr::kable(glance_lm, col.names = c("R-Squared"), digits = 3)

lm_fst_tidy <- tidy(lm_fst)

knitr::kable(lm_fst_tidy, col.names = c(
  "Term",
  "Estimate",
  "Standard Error",
  "Statistic",
  "P-Value"
  ), digits = 3)

```


```{r}

lm_reception <- lm(sequels_revenue ~ first_film_revenue + first_film_imdb, data = first_second_third)

glance_lm_imdb <- glance(lm_reception)|>
  select(adj.r.squared)

knitr::kable(glance_lm_imdb, col.names = c("Adjusted R-Squared"), digits = 3)

lm_reception_tidy <- tidy(lm_reception)

knitr::kable(lm_reception_tidy, col.names = c(
  "Term",
  "Estimate",
  "Standard Error",
  "Statistic",
  "P-Value"
  ), digits = 3)

```




```{r}

lm_reception2 <- lm(sequels_revenue ~ first_film_revenue + first_film_imdb + sequels_imdb, data = first_second_third)

glance_lm_imdb2 <- glance(lm_reception2)|>
  select(adj.r.squared)

knitr::kable(glance_lm_imdb2, col.names = c("Adjusted R-Squared"), digits = 3)

lm_reception2_tidy <- tidy(lm_reception2)

knitr::kable(lm_reception2_tidy, col.names = c(
  "Term",
  "Estimate",
  "Standard Error",
  "Statistic",
  "P-Value"
  ), digits = 3)


```

