---
title: "Gov 50 Final Project"
author: Jack Wyss
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---

```{r setup}
library(broom)
library(infer)
library(modelsummary)
library(moderndive)
library(skimr)
library(tidyverse)
```



## 1. Introduction

How do movie sequels compare to the original, in terms of box office return and audience reception? Does a better original film make for better sequels, or worse? Many people criticize the production of movie sequels, and yet studios keep making them. I hypothesize that a sequel will perform better if the first film in the series is more critically successful. This data is worth analyzing because movies and sequels are a continually relevant part of global cultural consumption and inform how we think about politics, news, and the arts.

If box office return of the original and average box office return of sequels are positively correlated, then there is evidence for part of my hypothesis. If the IMDb score of the original and the average IMDb score of its sequels are positively correlated, there is evidence for the other part of my hypothesis. If there is a negative correlation or no correlation for either, my hypothesis will be weakened. I will focus only on the first, second, and third films in a series for convenience, but there are still interesting things to be discovered in the data despite this limitation.

## 2. Data

There are 869 movies in this data set, which I formed by using left_join to combine two data sets:

  A) a data set containing information about movie series, including IMDb scores and the order of a film within a series
  B) a data set containing information about individual movies, including box office revenue, budget, and IMDb score. Unknown budgets or revenues are represented with a value of zero.
  
I created a variable (first_film_gt7, for "greater than seven") that determines whether a film was generally well-received (IMDb score >= 7) or poorly-received (IMDb score < 7), which I will use to estimate the average effect on sequel performance of the first film in the series having good or poor ratings. This is a cross-sectional research design for an observational study.

The data notes what franchise a film belongs to and the order in which it was released. For example, the original "Halloween" movie is given a value in the "Order" column of 1. The explanatory variables in this case are the box office revenue of the original, measured in dollars, and the IMDb score of the original, measured on a scale of 1 to 10 (with 10 being the best). The response variables are the box office revenue and IMDb scores of the original film's sequels. Since some sequels might be better-received than others, I will take the average gross box office return and IMDb score of the entire franchise after the first film. **Note that for movies released before IMDb existed the score is retroactive and may reflect modern consumer tastes rather than the tastes of people who went to see the movie in theaters.**


```{r}

movies <- read_csv("movies_metadata.csv")|>
  select(budget, overview, title, revenue, release_date, runtime)|>
  mutate(release_date = as.character(release_date))

sequels <- read_csv("Sequels/priyankad0993-sequels/Movies - Cleaned.csv")

movies_sequels <- sequels|>
  left_join(movies, by = c("Title" = "title", "Runtime (mins)" = "runtime"))|>
    select(Title, `Movie Series`, `IMDb Rating`, Order, budget, overview, revenue, `Runtime (mins)`)|>
  mutate(budget = as.integer(budget),
         revenue = as.integer(revenue),
    Order = as.character(Order),
    reception = if_else(`IMDb Rating` >= 7, "Good", "Poor"))|>
  rename(series = `Movie Series`,
         rating = `IMDb Rating`)

mean(movies_sequels$rating, na.rm = TRUE)

```

## 3. Analysis

**a) Average Number of Movies in a Series**


In this data set, movies are grouped by series, and each series has a certain number of movies. Some series have many sequels, and others are "one-and-done," so to speak; there is only one movie in the series. The average number of movies in series that have sequels is about 2.88. You can see on the histogram below that many series have two or three movies, whereas some have 10 or more. For the purposes of this project, I will be focusing on the first, second, and third films in a series.

```{r}

numbermovies <- movies_sequels|>
  group_by(series)|>
  summarize(number_of_movies = n())|>
  filter(number_of_movies > 1)

  mean(numbermovies$number_of_movies)
```
```{r}

numbermovies_plot <- numbermovies|>
  ggplot(mapping = aes(x = number_of_movies))+
  geom_bar()+
  labs(
    title = "Number of Movies by Franchise",
    x = "Number of Movies in the Franchise",
    y = "Count",
    caption = "Note: Excludes franchises with only one movie in the dataset."
  )+
  geom_vline(xintercept = mean(numbermovies$number_of_movies), color = "purple", size = 1.25)

numbermovies_plot

```

**b) Box Office Return (General)**

There is a strong positive correlation (0.7784) between a movie's budget and its box office return. As a movie's budget increases, it will generally make more money at the box office. Note that I filtered both the budget and revenue of a film to be greater than $10,000 to avoid capturing movies that did not get a theatrical release, or whose budgets are unknown.

```{r}

revenueplot <- movies_sequels|>
  drop_na()|>
  filter(budget > 10000 | revenue > 10000)|>
  ggplot(mapping = aes(x = budget, y = revenue))+
  geom_point()+
  geom_smooth(method = "lm", se = FALSE, color = "purple")+
  labs(title = "Budget vs Box Office Revenue",
       x = "Budget",
       y = "Revenue"
       )+
  scale_x_log10()+
  scale_y_log10()
  
revenueplot

cor(movies_sequels$budget, movies_sequels$revenue, use = "pairwise")
```

```{r}


movies_sequels <- movies_sequels|>
  mutate(sequel_category = case_when(
    Order == 1 ~ "First",
    Order == 2 ~ "Second",
    Order == 3 ~ "Third",
    Order == 4 ~ "Fourth",
    Order == 5 ~ "Fifth",
    Order == 6 ~ "Sixth",
    Order == 7 ~ "Seventh",
    Order == 8 ~ "Eighth",
    Order >= 9 ~ "Ninth +"),
    is.sequel = if_else(Order == 1, "No", "Yes")
  )

has_sequels_tibble <- movies_sequels|>
  group_by(series)|>
  summarize(number_of_movies = n())|>
  mutate(series.has.sequels = if_else(number_of_movies > 1, "Yes", "No"))

movies_sequels_new <- movies_sequels|>
inner_join(has_sequels_tibble, by = "series")

```

**c) ATE of First Film Reception on Sequel Revenue**

On average, sequel films whose original movie had an IMDb score greater than or equal to 7 had revenues that were 88% higher than sequel films whose original movie had an IMDb score less than 7. I chose 7 as the threshold for this test because the median first film IMDb score in the first_second_third data frame is 6.9.

```{r}

firstfilm <- movies_sequels_new|>
  drop_na()|>
  filter(series.has.sequels == "Yes")|>
  group_by(series, Order)|>
  summarize(first_film_revenue = revenue,
            first_film_imdb = rating,
            first_film_budget = budget,
            first_film_gt7 = if_else(first_film_imdb >= 7, "Yes", "No"))|>
  filter(Order == 1)

sequelfilms <- movies_sequels_new|>
  drop_na()|>
  filter(series.has.sequels == "Yes")|>
  group_by(series, Order)|>
  summarize(sequels_revenue = revenue,
            sequels_imdb = rating,
            sequels_budget = budget,
            sequels_gt7 = if_else(sequels_imdb >= 7, "Yes", "No"))|>
  filter(Order == 2 | Order == 3)

first_second_third <- firstfilm|>
  inner_join(sequelfilms, by = c("series"))|>
  rename(Order = Order.y)|>
  filter(first_film_revenue > 10000 & sequels_revenue > 10000)

head(first_second_third)

ate_good <- first_second_third|>
  group_by(first_film_gt7)|>
  summarize(mean_srevenue = mean(sequels_revenue, na.rm = TRUE))|>
  pivot_wider(names_from = first_film_gt7, values_from = mean_srevenue)|>
  mutate(ate = (Yes-No),
         ate_pct = (Yes-No) / No)

knitr::kable(ate_good, col.names = c("No", "Yes", "ATE", "ATE (Percent)"), digits = 3)

```

The p-value for this average treatment effect is nearly zero, which is less than an alpha-level of 0.1, 0.05, and 0.01. I reject the null hypothesis that sequel films whose first film was well-received do not have a difference in revenue from sequel films whose first film was poorly-received. This means there is evidence for my original hypothesis, which states that well-received original films produce more successful sequels.

```{r}

observed_ate <- first_second_third|>
  specify(formula = sequels_revenue ~ first_film_gt7)|>
  calculate(stat = "diff in means", order = c("Yes", "No"))

ATEtest <-  first_second_third|>
  specify(formula = sequels_revenue ~ first_film_gt7)|>
  hypothesize(null = "independence")|>
  generate(reps = 5000, type = "permute")|>
  calculate(stat = "diff in means", order = c("Yes", "No"))

visualize(ATEtest)+
  shade_p_value(obs_stat = observed_ate, direction = "both")+
  labs(x = "Average Treatment Effect (in Dollars)",
       y = "Count")
  
ATE_P <- ATEtest|>
  get_p_value(obs_stat = observed_ate, direction = "both")

knitr::kable(ATE_P, col.names = "P-Value", digits = 3)
```

**d) ATE of First Film Reception on Second and Third Film Revenue**

I thought it might be the case that the treatment would affect films differently depending on their position in a franchise. On average, the revenues of a second film whose original movie had an IMDb score greater than or equal to 7 were 89.5% higher than the revenues of a second film whose original movie had an IMDb score less than 7. On average, the revenues of a third film whose original movie had an IMDb score greater than or equal to 7 were 63.9% higher than the revenues of a second film whose original movie had an IMDb score less than 7. A well-received original film appears to affect the revenues of second films more than the revenues of third films.

```{r}

ate_good <- first_second_third|>
  group_by(first_film_gt7, Order)|>
  summarize(mean_srevenue = mean(sequels_revenue, na.rm = TRUE))|>
  pivot_wider(names_from = first_film_gt7, values_from = mean_srevenue)|>
  mutate(ate = (Yes-No),
         ate_pct = (Yes-No) / No)

knitr::kable(ate_good, col.names = c("Order", "No", "Yes", "ATE", "ATE (Percent)"), digits = 3)

```

**e) Regression**

I was worried about confounding variables in my data set. From the following regression, we can see that the R-squared value for this data is *very* low at only 0.068. This means only 6.8% of the variation in the data is explained by the regression model which includes only whether the first film in a series has an IMDb score of 7 or greater.

```{r}

lm_fst <- lm(sequels_revenue ~ first_film_gt7, data = first_second_third)

glance_lm_fst <- glance(lm_fst)|>
  select(r.squared)

knitr::kable(glance_lm_fst, col.names = c("R-Squared"), digits = 3)

lm_fst_tidy <- tidy(lm_fst)

knitr::kable(lm_fst_tidy, col.names = c(
  "Term",
  "Estimate",
  "Standard Error",
  "Statistic",
  "P-Value"
  ), digits = 3)

```

Including sequels_gt7, which tells whether the sequel film has an IMDb score greater than or equal to 7, increases the adjusted R-squared value to 0.254. This means that 25.4% of the variation in the data is explained by the model. Additionally, the inclusion of sequels_gt7 in the model causes the p-value of the original ATE to increase to 0.079. *This means we can no longer reject the null hypothesis stating that there is no difference in revenues for sequel films whose original films have IMDb scores greater than or equal to 7 and those whose original films have IMDb scores less than 7 at an alpha-level of 0.05, suggesting that sequels_gt7 is a confounding variable.* Although the adjusted R-squared value for this model is still small, I would still say that my hypothesis has been weakened.

```{r}

lm_multiple <- lm(sequels_revenue ~ first_film_gt7 + sequels_gt7, data = first_second_third)

glance_lm_multiple <- glance(lm_multiple)|>
  select(adj.r.squared)

knitr::kable(glance_lm_multiple, col.names = c("Adjusted R-Squared"), digits = 3)

lm_multiple_tidy <- tidy(lm_multiple)

knitr::kable(lm_multiple_tidy, col.names = c(
  "Term",
  "Estimate",
  "Standard Error",
  "Statistic",
  "P-Value"
  ), digits = 3)

```



```{r}

modelsummary(list(lm_fst, lm_multiple),
             gof_map = c("nobs", "r.squared","adj.r.squared"))

```


